An Image-based method for phase estimation, gating and temporal super-resolution of \\cardiac ultrasound

### 2017-02-23 paper155.tex
Made QRS peak identification table look better
Playing with periodogram figue
Updated paper

\begin{abstract}
Ultrasound is emerging as an increasingly effective tool for rapid non-invasive assessment of cardiac structure and function. Knowledge of the phase or location of each video frame within the cardiac and/or respiratory cycles is essential in many applications. In this paper, we present a novel method for estimation of instantaneous cardiac and respiratory phases directly from cardiac ultrasound videos. The method transforms the complex high-dimensional image sequence into a univariate time series with the same periodicity characteristics, decouples the periodic sources of beating heart and respiratory motion using a trend extraction technique, and estimates the cardiac and respiratory phases using the Hilbert transform. We also present a robust non-parametric regression technique for gating out respiratory frames and a novel kernel regression model that reconstructs images at any cardiac phase and facilitates temporal super-resolution. We validate our methods using cardiac ultrasound or echocardiography videos and ECG recordings of 6 mice.
%\keywords{Cardiac, Ultrasound, Echocardiography, Phase estimation, Gating, Temporal Super-resolution}
\vspace{-0.5cm}
\end{abstract}
%
\section{Introduction}
\label{sec:intro}
%
Cardiovascular disease is the leading cause of death worldwide and ultrasound is emerging as an increasingly effective tool for rapid non-invasive assessment of cardiac structure and function. Cardiac ultrasound videos consist of two kinds of periodic motion, one corresponding to beating heart motion and the other corresponding to respiratory motion. Knowledge of the phase or location of each video frame within the cardiac and/or respiratory cycles is essential in a variety of applications such as gating\cite{VonBirgelen1997}, quiescence detection\cite{Wick2013}, 3D reconstruction\cite{Wachinger2012}, and temporal super-resolution\cite{Cherin2006}. Typically, the position within cardiac cycle is tracked through ECG data and position within respiratory cycle is tracked by motion of markers placed on the subject's body\cite{Khamene2004}.

	In this paper, we present a novel method that estimates the instantaneous cardiac and respiratory phases directly from the cardiac ultrasound video. We also present a robust non-parametric regression technique for gating out respiratory frames and a novel kernel regression to reconstruct images at any cardiac phase to facilitate temporal super-resolution. Previous work on the estimation of cardiac and/or respiratory phases directly from ultrasound videos is very limited. In~\cite{Karadayi2006}, the authors compute a signal measuring the x- or y-coordinate of the center of mass of each frame, use bandpass filtering to remove components outside typical cardiac frequency range, find the dominant frequency and apply matched sine/cosine filters to estimate the instantaneous phase. In~\cite{Sundar2009}, the authors compute a signal measuring phase correlation between consecutive frames, use a bandpass filter centered at 1Hz to recover the cardiac signal, and use a low pass filter to recover the respiratory signal.

@@

In this section, we present the theory underlying the proposed methods. Specifically, we begin by describing our method for the estimation of instantaneous cardiac and respiratory phases in Section \ref{sec:method:phase_estimation}. We then present a robust method that uses these phase estimates to gate out video frames with significant respiratory motion in Section \ref{sec:method:gating}. Finally, in Section \ref{sec:method:super_resolution}, we present a kernel regression model for reconstructing images at any cardiac phase and facilitate temporal super-resolution.

@@

While there have been numerous efforts for the estimation of instantaneous phase and/or frequency in periodic univariate time series data\cite{Boashash1992,Lu2013,Luo2003}, there are not many methods that tackle this problem in a multivariate setting such as the case of periodic cardiac ultrasound videos wherein several thousands of variables (pixel intensities) are involved. Our strategy was to find a way to transforming this complex multi-variate problem into a univariate one and take advantage of existing univariate methods to solve the problem.

@@

	In the second step, we learn a robust mapping $L(\phi) : \left [  0, 1\right ) \to R$ from cardiac phase to the residual/heart-beat component of the frame similarity signal by fitting a robust non-parametric regression model called Locally weighted regression (LOWESS) to the dataset $\left \{ \left(\hat{\phi}_{heart}(t), \hat{r}_{heart}(t) \right) \mid t \notin F_{cutoff}  \right \}$ containing the pair of the cardiac phase $\hat{\phi}_{heart}(t)$ and residual/heart-beat $\hat{r}_{heart}(t)$ signal values of all frames that do not belong to the set of frames $F_{cutoff}$ discarded in the first step above. Next, we compute a robust estimate of the standard deviation/dispersion $\hat{\sigma}_{L}$ of non-respiratory frames around the LOWESS fit based on the median absolute value of the difference between the LOWESS fit $L( \hat{\phi}_{heart}(t) )$ and residual/heart-beat signal $\hat{r}_{heart}(t)$ for all frames. Lastly, we gate out all frames $F_{resp} = \left \{ t \; \big\lvert \; \lvert L( \hat{\phi}_{heart}(t) ) - \hat{r}_{heart}(t)  \rvert   > k \times \hat{\sigma}_{L}  \right \}$ for which the absolute  difference between the LOWESS fit and heart-beat signal value is greater than $k = 2.0$ (95\% confidence interval) times the standard deviation $\hat{\sigma}_{L}$.	
%
\begin{IonFigT}
\centering
%
\includegraphics[width=4.5in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/phaseordered.png}
\ionbox{1.2in}\ionbox{1.1in}\ionbox{1.1in}\ionbox{1.1in}\\
%
\caption{Visualization of m-mode frames ordered by cardiac phase: (a) One of the frames in the video overlaid with the M-mode line shown in the next three images, (b) M-mode frames in the order they appear in the input video, (c) M-mode frames ordered by cardiac phase, and (d) Non-respiratory m-mode frames ordered by cardiac phase.}
\label{fig:phase_ordering}
\end{IonFigT}
%
\vspace{-0.5cm}
\subsection{Model to generate images by cardiac phase for super-resolution}
\label{sec:method:super_resolution}
%
In this section, we present a kernel regression model to reconstruct the image at any cardiac phase. This model can then be used to generate a single-cycle video representative of the subject's heart-beat at a higher temporal resolution.

Given a cardiac ultrasound video $I(t) : \{1, ..., N\} \to R^m$ of $N$ images with $m$ pixels each, we extract the residual/heart-beat component $\hat{r}_{heart}(t)$ of the frame similarity signal, estimate the instantaneous cardiac phase $\hat{\phi}_{heart}(t)$ of each frame, and compute the robust LOWESS fit $L(\phi) : \left [  0, 1\right ) \to R$ that maps cardiac phase to the heart-beat signal as  described in Sections~\ref{sec:method:phase_estimation} and ~\ref{sec:method:gating}. We then use Nadarya-Watson kernel regression\cite{Bishop2006} to learn a phase-parametrized image manifold in the form of a function $M(\phi): [0, 1) \to R^m $ that maps any cardiac phase $\phi$ to an image using kernel-weighted local average as follows:
\begin{equation}
M(\phi) = \frac{\sum_{t = 1}^{N} K \left( \phi, \hat{\phi}_{heart}(t) \right) I(t)}{\sum_{t = 1}^{n} K \left( \phi, \hat{\phi}_{heart}(t) \right)} 
\end{equation}
where kernel $K\left( \phi, \hat{\phi}_{heart}(t) \right) = exp\left \{ -\frac{ \mid \phi - \hat{\phi}_{heart}(t) \mid^2}{2  \sigma^2_\phi} \right \} \times exp\left \{ -\frac{ \mid L(\phi) - \hat{r}_{heart}(t) \mid^2}{2  \sigma^2_{L}} \right \}$ is defined as the product of two radial-basis function (RBF) kernels. The first RBF kernel  weighs images inversely proportional to their distance (accounting for periodicity) in the cardiac phase space. Its bandwidth $\sigma_\phi$ is set equal to a constant $k_\phi = 0.4$ times the median difference in cardiac phase between consecutive frames of the given image sequence. The second RBF kernel weighs images inversely proportional to their deviation from the LOWESS prediction $L(\phi)$ in the residual/heart-beat signal space. Its bandwidth $\sigma_{L}$ is set equal to a constant $k_L = 2.0$ times the robust estimate of standard deviation/dispersion $\hat{\sigma}_{L}$ of non-respiratory frames around the LOWESS fit as described in Section~\ref{sec:method:gating}. A single-cycle video representative of the subject's heart-beat can now be reconstructed by sampling the manifold $M(\phi)$ at any desired resolution. 
\vspace{-0.5cm}
\section{Results}
\label{sec:results}
%
We use cardiac ultrasound videos and associated ECG recordings of 6 mice to validate our methods. The ultrasound videos were acquired using the VisualSonics Vevo 2100 scanner at 233 frames per second. Each video consists of approximately 300 frames, 11 cardiac cycles, and 2 respiratory cycles.

@@

\begin{minipage}[b]{0.45\linewidth}

@@

\begin{tabular}{|c|c|c|c|}

@@

\begin{tabular}[c]{@{}c@{}}VID\\ $\;$ \end{tabular} & \begin{tabular}[c]{@{}c@{}}Correlation \\ $R^2$ \end{tabular} & \begin{tabular}[c]{@{}c@{}}QRS peak prediction\\ $mean \pm std \; error$\end{tabular} \\ \hline
1 & 0.9995 & 1.55 $\pm$ 0.89 \\ \hline
2 & 0.9995 & 1.36 $\pm$ 1.15 \\ \hline
3 & 0.9996 & 1.45 $\pm$ 0.89 \\ \hline
4 & 0.9998 & 0.73 $\pm$ 0.62 \\ \hline
5 & 0.9990 & 2.09 $\pm$ 1.78 \\ \hline
6 & 0.9992 & 1.40 $\pm$ 1.74 \\ \hline

>>>>

\begin{abstract}
Ultrasound is emerging as an increasingly effective tool for rapid non-invasive assessment of cardiac structure and function. Knowledge of the phase or location of each video frame within the cardiac and/or respiratory cycles is essential in many applications. In this paper, we present a novel method for estimation of instantaneous cardiac and respiratory phases directly from cardiac ultrasound videos. The method transforms the complex high-dimensional image sequence into a univariate time series with the same periodicity characteristics, decouples the periodic sources of beating heart and respiratory motion using a trend extraction technique, and estimates the cardiac and respiratory phases using the Hilbert transform. We also present a robust non-parametric regression technique for gating out respiratory frames and a kernel regression model that reconstructs images at any cardiac phase and facilitates temporal super-resolution. We validate our methods using cardiac ultrasound or echocardiography videos and ECG recordings of 6 mice.
%\keywords{Cardiac, Ultrasound, Echocardiography, Phase estimation, Gating, Temporal Super-resolution}
\vspace{-0.5cm}
\end{abstract}
%
\section{Introduction}
\label{sec:intro}
%
Cardiovascular disease is the leading cause of death worldwide and ultrasound is emerging as an increasingly effective tool for rapid non-invasive assessment of cardiac structure and function. Cardiac ultrasound videos consist of two kinds of periodic motion, one corresponding to beating heart motion and the other corresponding to respiratory motion. Knowledge of the phase or location of each video frame within the cardiac and/or respiratory cycles is essential in a variety of applications such as gating\cite{VonBirgelen1997}, quiescence detection~\cite{Wick2013}, 3D reconstruction~\cite{Wachinger2012}, and temporal super-resolution~\cite{Cherin2006}. Typically, the position within cardiac cycle is tracked through ECG data and position within respiratory cycle is tracked by motion of markers placed on the subject's body\cite{Khamene2004}.

	In this paper, we present a novel method that estimates the instantaneous cardiac and respiratory phases directly from the cardiac ultrasound video. We also present a robust non-parametric regression technique for gating out respiratory frames and a novel kernel regression to reconstruct images at any cardiac phase to facilitate temporal super-resolution. Previous work on the estimation of cardiac and/or respiratory phases directly from ultrasound videos is very limited. In~\cite{Karadayi2006}, the authors compute a signal measuring the x- or y-coordinate of the center of mass of each frame, use bandpass filtering to remove components outside typical cardiac frequency range, find the dominant frequency and apply matched sine/cosine filters to estimate the instantaneous phase. In~\cite{Sundar2009}, the authors compute a signal measuring phase correlation between consecutive frames, use a bandpass filter to recover the cardiac signal and a low pass filter to recover the respiratory signal. 

@@

In this section, we present the theory underlying the proposed methods. Specifically, we begin by describing our method for estimation of instantaneous cardiac and respiratory phases in Section \ref{sec:method:phase_estimation}. We then present a robust method to gate out video frames with significant respiratory motion in Section \ref{sec:method:gating}. Finally, in Section \ref{sec:method:super_resolution}, we present a kernel regression model for reconstructing images at any cardiac phase and facilitate temporal super-resolution.

@@

While there have been numerous efforts for the estimation of instantaneous phase and/or frequency in periodic univariate time series data~\cite{Boashash1992,Lu2013,Luo2003}, there are not many methods that tackle this problem in a multivariate setting such as the case of periodic cardiac ultrasound videos wherein several thousands of variables (pixel intensities) are involved. Our strategy is to transform this complex multi-variate problem into a univariate one and take advantage of existing methods to solve the problem.

@@

	In the second step, we learn a robust mapping $L(\phi) : \left [  0, 1\right ) \to R$ from cardiac phase to the residual/heart-beat component of the frame similarity signal by fitting a robust non-parametric regression model called Locally weighted regression (LOWESS) to the dataset $\left \{ \left(\hat{\phi}_{heart}(t), \hat{r}_{heart}(t) \right) \mid t \notin F_{cutoff}  \right \}$ containing the pair of the cardiac phase $\hat{\phi}_{heart}(t)$ and residual/heart-beat $\hat{r}_{heart}(t)$ signal values of all frames that do not belong to the set of frames $F_{cutoff}$ discarded in the first step above. Next, we compute a robust estimate of the standard deviation/dispersion $\hat{\sigma}_{L}$ of non-respiratory frames around the LOWESS fit based on the median absolute value of the difference between the LOWESS fit $L( \hat{\phi}_{heart}(t) )$ and residual/heart-beat signal $\hat{r}_{heart}(t)$ for all frames. Lastly, we gate out all frames $F_{resp} = \left \{ t \; \big\lvert \; \lvert \hat{r}_{heart}(t) - L( \hat{\phi}_{heart}(t) ) \rvert   > k \times \hat{\sigma}_{L}  \right \}$ for which the absolute  difference between the LOWESS fit and heart-beat signal value is greater than $k = 2.0$ (95\% confidence interval) times the standard deviation $\hat{\sigma}_{L}$.	
%
\vspace{-0.5cm}
\subsection{Model to generate images by cardiac phase for super-resolution}
\label{sec:method:super_resolution}
%
In this section, we present a kernel regression model to reconstruct the image at any cardiac phase. This model can then be used to generate a single-cycle video representative of the subject's heart-beat at a higher temporal resolution.

Given a cardiac ultrasound video $I(t) : \{1, ..., N\} \to R^m$ of $N$ images with $m$ pixels each, we extract the residual/heart-beat component $\hat{r}_{heart}(t)$ of the frame similarity signal, estimate the instantaneous cardiac phase $\hat{\phi}_{heart}(t)$ of each frame, and compute the robust LOWESS fit $L(\phi) : \left [  0, 1\right ) \to R$ that maps cardiac phase to the heart-beat signal as  described in Sections~\ref{sec:method:phase_estimation} and ~\ref{sec:method:gating}. We then use Nadarya-Watson kernel regression\cite{Bishop2006} to learn a phase-parametrized image manifold in the form of a function $M(\phi): [0, 1) \to R^m $ that maps any cardiac phase $\phi$ to an image using kernel-weighted local average as follows:
\begin{equation}
M(\phi) = \frac{\sum_{t = 1}^{N} K \left( \phi, \hat{\phi}_{heart}(t) \right) I(t)}{\sum_{t = 1}^{n} K \left( \phi, \hat{\phi}_{heart}(t) \right)} 
\end{equation}
where kernel $K\left( \phi, \hat{\phi}_{heart}(t) \right) = exp\left \{ -\frac{ \mid \phi - \hat{\phi}_{heart}(t) \mid^2}{2  \sigma^2_\phi} \right \} \times exp\left \{ -\frac{ \mid L(\phi) - \hat{r}_{heart}(t) \mid^2}{2  \sigma^2_{L}} \right \}$ is defined as the product of two radial-basis function (RBF) kernels. The first RBF kernel  weighs images inversely proportional to their distance (accounting for periodicity) in the cardiac phase space. Its bandwidth $\sigma_\phi$ is set equal to a constant $k_\phi = 0.4$ times the median difference in cardiac phase between consecutive frames of the given image sequence. The second RBF kernel weighs images inversely proportional to their deviation from the LOWESS prediction $L(\phi)$ in the residual/heart-beat signal space. Its bandwidth $\sigma_{L}$ is set equal to a constant $k_L = 2.0$ times the robust estimate of standard deviation/dispersion $\hat{\sigma}_{L}$ of non-respiratory frames around the LOWESS fit as described in Section~\ref{sec:method:gating}. A single-cycle video representative of the subject's heart-beat can now be reconstructed by sampling the manifold $M(\phi)$ at any desired resolution. 
%
%
\begin{IonFigT}
\centering
%
\includegraphics[width=4.5in]{figures/decoded/2015-07-27-10-36-06_2015-07-15-16-56-16_1.raw.bmode/phaseordered.png}
\ionbox{1.2in}\ionbox{1.1in}\ionbox{1.1in}\ionbox{1.1in}\\
%
\caption{Visualization of m-mode frames ordered by cardiac phase: (a) One of the frames in the video overlaid with the M-mode line shown in the next three images, (b) M-mode frames in the order they appear in the input video, (c) M-mode frames ordered by cardiac phase, and (d) Non-respiratory m-mode frames ordered by cardiac phase.}
\label{fig:phase_ordering}
\end{IonFigT}
%
%
\vspace{-0.5cm}
\section{Results}
\label{sec:results}
%
We use cardiac ultrasound videos and associated ECG recordings of 6 mice to validate our methods. The ultrasound videos were acquired using the VisualSonics Vevo 2100 scanner at 233 frames per second. Each video consists of approximately 300 frames, 11 cardiac cycles, and 2 respiratory cycles.

@@

\begin{minipage}[t]{0.45\linewidth}

@@

@@

\multicolumn{1}{|l|}{} & \multicolumn{2}{l|}{QRS peak frame identification} \\ \hline
VID & $R^2$ & $mean \pm std \; error$ \\ \hline
1 & \; 0.9995 \; & 1.55 $\pm$ 0.89 \\ \hline
2 & \; 0.9995 \; & 1.36 $\pm$ 1.15 \\ \hline
3 & \; 0.9996 \; & 1.45 $\pm$ 0.89 \\ \hline
4 & \; 0.9998 \; & 0.73 $\pm$ 0.62 \\ \hline
5 & \; 0.9990 \; & 2.09 $\pm$ 1.78 \\ \hline
6 & \; 0.9992 \; & 1.40 $\pm$ 1.74 \\ \hline
